---
title: "Batch correction using Seurat"
author: "Britta Velten"
date: "06/01/2021"
output: BiocStyle::html_document
---

Following the [tutorial provided by Seurat (v3)](https://satijalab.org/seurat/v3.1/integration.html) we will integrate human pancreatic islet cells profiled by different technologies.

# Getting the data
```{r, warning=FALSE, message=FALSE}
library(Seurat)
library(SeuratData)
```

```{r}
# InstallData("panc8")
data("panc8")
pancreas.list <- SplitObject(panc8, split.by = "tech")
pancreas.list <- pancreas.list[c("celseq", "celseq2", "fluidigmc1", "smartseq2")]
```

# Data visualization before integration
```{r}
panc8 <- NormalizeData(panc8)
panc8 <- FindVariableFeatures(panc8)
panc8 <- ScaleData(panc8)
panc8 <- RunPCA(panc8, npcs = 30, verbose = FALSE)
panc8 <- RunUMAP(panc8, reduction = "pca", dims = 1:30)
DimPlot(panc8, group.by = "tech", reduction = "umap")
DimPlot(panc8, group.by ="celltype", reduction = "umap")
```

# Data normalization per batch
```{r}
for (i in 1:length(pancreas.list)) {
    pancreas.list[[i]] <- NormalizeData(pancreas.list[[i]], verbose = FALSE)
    pancreas.list[[i]] <- FindVariableFeatures(pancreas.list[[i]], selection.method = "vst", 
        nfeatures = 2000, verbose = FALSE)
}
```

# Integration of batches
```{r}
reference.list <- pancreas.list[c("celseq", "celseq2", "smartseq2")]
pancreas.anchors <- FindIntegrationAnchors(object.list = reference.list, dims = 1:30)
```

```{r}
pancreas.integrated <- IntegrateData(anchorset = pancreas.anchors, dims = 1:30)
```

# Data vizualization after integration

```{r, fig.width=10}
library(ggplot2)
library(cowplot)
library(patchwork)
# switch to integrated assay. The variable features of this assay are automatically
# set during IntegrateData
DefaultAssay(pancreas.integrated) <- "integrated"

# Run the standard workflow for visualization and clustering
pancreas.integrated <- ScaleData(pancreas.integrated, verbose = FALSE)
pancreas.integrated <- RunPCA(pancreas.integrated, npcs = 30, verbose = FALSE)
pancreas.integrated <- RunUMAP(pancreas.integrated, reduction = "pca", dims = 1:30)
p1 <- DimPlot(pancreas.integrated, reduction = "umap", group.by = "tech")
p2 <- DimPlot(pancreas.integrated, reduction = "umap", group.by = "celltype", label = TRUE, 
    repel = TRUE) + NoLegend()
p1 + p2
```


For further details have also a look at the more extended Seurat (v4) tutorial including various downstream analyses on the integrated data: https://satijalab.org/seurat/articles/integration_introduction.html.

